---
title: "Analysis of Heart Disease Dataset"
author: "Emma Juan Salazar (emma.juan01@estudiant.upf.edu) and Gary etc."
date: "Last update: `r format(Sys.time(), '%d %B, %Y')`"      
output:
  html_document:
    toc: true
    fig_caption: true
  html_notebook: 
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Analysis of the Heart Disease Dataset

Load the data from [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_dataset.csv), and the description is [here](https://raw.githubusercontent.com/jpinero/DMI_2021/main/datasets/heart_disease_description.txt). The original dataset comes from [here](https://archive.ics.uci.edu/ml/datasets/Heart+Disease) and corresponds to the [processed cleveland data](https://archive.ics.uci.edu/ml/machine-learning-databases/heart-disease/processed.cleveland.data)

## Perform an EDA on the dataset

## 1. Getting to know the dataset

First, we import the necessary libraries

```{r}
library(ggplot2)
library(DataExplorer)
library(mlbench)
library(GGally)
library("tidyverse")
library("vtable")
library("SmartEDA")
library("corrplot")
```

## The features on our dataset are the following

1.  age (Age in years)

2.  sex : (1 = male, 0 = female)

3.  cp (Chest Pain Type): [ 0: asymptomatic, 1: atypical angina, 2: non-anginal pain, 3: typical angina]

4.  trestbps (Resting Blood Pressure in mm/hg )

5.  chol (Serum Cholesterol in mg/dl)

6.  fps (Fasting Blood Sugar \> 120 mg/dl): [0 = no, 1 = yes]

7.  restecg (Resting ECG): [0: showing probable or definite left ventricular hypertrophy by Estes\' criteria, 1: normal, 2: having ST-T wave abnormality]

8.  thalach (maximum heart rate achieved)

9.  exang (Exercise Induced Angina): [1 = yes, 0 = no]

10. oldpeak (ST depression induced by exercise relative to rest)

11. slope (the slope of the peak exercise ST segment): [0: downsloping; 1: flat; 2: upsloping]

12. ca [number of major vessels (0--3)

13. thal : [1 = normal, 2 = fixed defect, 3 = reversible defect]

14. target: [0 = disease, 1 = no disease]

We import the data, and we name the columns.

```{r}
heart_data = read.csv("data/heart_disease_dataset.csv", sep=" ")

colnames(heart_data) <- c("Age", "Sex", "ChestPain", "RBP", "Cholesterol", "FastingBloodSugar", "RestingECResults", "MaxHeartRate", "ExInducedAngina", "Oldpeak", "Slope", "NbMajorVessels", "Thalassemia", "Num", "PatetientID")

# We use dim() to obtain the number of rows and columns on our dataset

dim(heart_data)
```

We check the type of the variables

```{r}
str(heart_data)
```

We see that the features "NbMajorVessels" and "Thalassemia" are characters. We have to convert them to numerical types.

```{r}
transform(heart_data, "Thalassemia" = as.numeric("Thalassemia"))
```

## 2. Data cleaning

```{r}
ExpData(data=heart_data,type=2, fun = c("mean", "median", "var"))
```

## Create visualizations in order to show which variables seem to be more associated with heart disease

```{r}

```

# 2 Difference in mortality rates in hospitalized COVID-19 patients

Using the supplementary material from the [Difference in mortality rates in hospitalized COVID-19 patients identified by cytokine profile clustering using a machine learning approach: An outcome prediction alternative](https://www.frontiersin.org/articles/10.3389/fmed.2022.987182/full), perform the following tasks

## Reproduce Figure 1 from the publication

```{r}

```

## Reproduce Figure 2 from the publication

but instead of representing the clusters in the annotation, represent the groups (G1 to G4)

```{r}

```

## Improve figure 2 of the publication

Add a second annotation with information of deathm and a third one with information of gender

```{r}

```

# session info {.unnumbered}

```{r, results='asis',  echo=FALSE, message=FALSE }
sessionInfo()
```
